Algebraic Tori in Unitary Groups

### 2014-03-23 Paper.tex
reorganize a bit
Reorganize the description of forms of Hermitian forms and forms of
Unitary groups.

For our purposes a unitary group is a simply connect group $G$ of type $A_n$.

@@

An element of $H^1(\GalKbK, \mu_2)$ corresponds to a quadratic \'etale extension $E$ of $K$.
More precisely it is the $n+1$ torsion of $T_{E,\sigma}$.
We have an exact sequence:
\[ \ldots \rightarrow H^1(\GalKbK,G) \rightarrow H^1(\GalKbK, G^{adj}) \rightarrow H^2(\GalKbK, \mu_{n+1}^\xi) \rightarrow \ldots. \]
A form of $G$ thus has an invariant in $H^2(\GalKbK, mu_{n+1}^\xi)$.

$H^2(\GalKbK, \mu_{n+1}^\xi)$ fits into the exact sequence:
\[ H^1(\GalKbK, T_{E,\sigma}) \rightarrow H^2(\GalKbK, mu_{n+1}^\xi) \rightarrow H^2(\GalKbK, T_{E,\sigma}) \overset{[n+1]}\rightarrow H^2(\GalKbK, T_{E,\sigma}). \]
We are thus interested in the cohomology of $T_{E,\sigma}$.

The exact sequence $1 \rightarrow T_{E,\sigma} \rightarrow T_E \rightarrow T_{E^\sigma} \rightarrow 1$ and Hilbert's Theorem 90 allow us to compute that 
$H^1(\GalKbK, T_{E,\sigma}) = k^\times/N_{E/k}(E^\times)$ and that $H^2(\GalKbK, T_{E,\sigma})$ is the kernel of the corestriction map from the Brauer group of $E$ to that of $E^\sigma$.
In particular we can now deduce that $H^2(\GalKbK, \mu_{n+1}^\xi)$ corresponds to element a pair of an element of $\delta\in k^\times/(k^\times)^nN_{E/k}(E^\times)$ and a degree $n$ central simple algebra over $E$ in the $n$-torsion of the Brauer group which is only non-trivial at the places of $E$ over $K$ which split, and for places which split the algebra over one factor is isomorphic to the opposite algebra of that over the other.

\begin{remark}
It should be pointed out that the group $K^\times/(K^\times)^nN_{E/K}(E^\times)$ is trivial if either $n$ is odd or if $E = K\times K$ is split.
\end{remark}

Finally, we must consider the group $H^1(\GalKbK,G)$. As $G$ is simply connect, this is only non-trivial at the real places of $K$.
In terms of real places, there are two cases to consider.
If $E=\C$, then $G$ is the automorphism group of a Hermitian space and $H^1(\GalKbK,G)$ classifies the Hermitian forms of the same discriminant as that defining $G$, as above, over $\RR$ this means it classifies the signature.
When $E=\RR\times\RR$ then the $G$ is the automorphism group of a free module.
 As there are no forms of free modules, $H^1(\GalKbK,G)$ is then trivial. 
TODO --- Find reference for this ----

@@

Given a a collection of invariants, $E$, $A$, $\delta$, and $(s_\nu,r_\nu)$ we can construct the group $G$ as follows.
By assumption $\sigma(A) \simeq A^{op}$ and thus $A$ can be given an involution $\sigma_A$ restricting to $\sigma$ on $E$.
Fix an element $J \in A^\sigma$ of reduced norm $\delta$ such that at each real place of $K$ which ramifies in $E$ the matrix for $J$ has $r$ positive eigenvalues and $s$ negative eigenvalues where $(r,s)$ is the signature at the real place. We may then define $G$ as:
\[ G(R) = \{ g \in (A\otimes_K R)^\times \mid gJ\sigma_A(g) = J \} \]
One may check that $G$ determines $E$, $A$, $\delta$ and the signatures uniquely when $n>1$.

\begin{remark}
In the case of $n=1$, we may perform the same constructions, however the group does not uniquely determine invariants $E$, $A$, $\delta$, and signatures.
Indeed, there is no outer automorphism group, so the choice of $E$ is no longer an invariant.
Exceptional isomorphisms of groups tell us that for $n=1$ the group $G$ is the spin group of a rank $3$ quadratic space.
In odd rank the discriminant does not impact the isomorphism type of the orthogonal group.
The discriminant of the quadratic space does still play a role in defining the Clifford algebra, in the same way it plays a role in defining a unitary group. The Witt invariant of the quadratic space is still an invariant and determines the algebra $A$.

>>>>

For our purposes a unitary group is a simply connect group $G$ of type $A_n$.
We may construct such a group as follows.
Fix the following:
\begin{itemize}
\item An quadratic \'etale algebra $E$ over $K$.
\item A central simple $E$-algebra $F$ of degree $n$ with anti-involution $\sigma$ restricting to the non-trivial involution of $E$.
\item An element $J\in F^\sigma$ of norm $\delta$ whose minimal polynomial at a real place $\nu$ of $K$ which ramifies in $E$ has $r_\nu$ positive roots and $s_\nu$ negative roots.
\end{itemize}
Define then $G$ to be the group scheme whose points over a $K$-algebra $R$ are given by:
\[ G(R) = \{ g\in (F\otimes_K R)^\times \mid gJ\sigma(g) = J \}. \]

We claim that all $G$ arrise this way, and that if $n>1$ then $G$ is uniquely determined by the invariants $(E,A,\delta, \left\vert r_\nu-s_\nu\right\vert)$ where $\delta \in K^\times/(K^\times)^nN_{E/K}(E^\times)$ subject only to the condition that $(-1)^{r_\nu-s_\nu}\delta$ is positive at all the real places of $K$ which ramify in $E$.

@@

An element of $H^1(\GalKbK, \mu_2)$ corresponds to the quadratic \'etale extension $E$ of $K$.

The center $\mu_{n+1}^\xi)$ of $G$ is the $n+1$ torsion of $T_{E,\sigma}$.
We have an exact sequence:
\[ \ldots \rightarrow H^1(\GalKbK,G) \rightarrow H^1(\GalKbK, G^{adj}) \rightarrow H^2(\GalKbK, \mu_{n+1}^\xi) \rightarrow \ldots. \]
A form of $G$ thus has an invariant in $H^2(\GalKbK, mu_{n+1}^\xi)$.

$H^2(\GalKbK, \mu_{n+1}^\xi)$ fits into the exact sequence:
\[ H^1(\GalKbK, T_{E,\sigma}) \rightarrow H^2(\GalKbK, mu_{n+1}^\xi) \rightarrow H^2(\GalKbK, T_{E,\sigma}) \overset{[n+1]}\rightarrow H^2(\GalKbK, T_{E,\sigma}). \]
We are thus interested in the cohomology of $T_{E,\sigma}$.
The exact sequence $1 \rightarrow T_{E,\sigma} \rightarrow T_E \rightarrow T_{E^\sigma} \rightarrow 1$ and Hilbert's Theorem 90 allow us to compute that 
$H^1(\GalKbK, T_{E,\sigma}) = k^\times/N_{E/k}(E^\times)$ and that $H^2(\GalKbK, T_{E,\sigma})$ is the kernel of the corestriction map from the Brauer group of $E$ to that of $E^\sigma$.
In particular we can now deduce that $H^2(\GalKbK, \mu_{n+1}^\xi)$ corresponds to element a pair of an element of $\delta\in k^\times/(k^\times)^nN_{E/k}(E^\times)$ and a degree $n$ central simple algebra $A$ over $E$ in the $n$-torsion of the Brauer group which is only non-trivial at the places of $E$ over $K$ which split, and for places which split the algebra over one factor is isomorphic to the opposite algebra of that over the other.

It is precisely these algebras $A$ which admit non-trivial involutions acting non-trivially on there center.
Indeed, in this setting we have that $A \otimes_{E,\sigma} E \simeq A^{op}$, that is if we twist the underlying $E$ algebra structure of $A$ then the resulting algebra is isomorphic to the opposite algebra of $A$ as an $E$-algebra. Let $\tau$ be the composition of the maps $A \rightarrow A \otimes_{E,\sigma} E \simeq A^{op} \rightarrow A$. This is an anti-involution on $A$, and as the map $A \rightarrow A \otimes_{E,\sigma} E $ is not a map of $E$-algebras (only of $K$-algebras) the map cannot act trivially on $E$. It must however preserve $E$ as this is the center of $A$. Thus $\tau$ restricts to $\sigma$ on $E$.

TODO - ref book of involutions re conjugacy of these involutions

\begin{remark}
It should be pointed out that the group $K^\times/(K^\times)^nN_{E/K}(E^\times)$ is trivial if either $n$ is odd or if $E = K\times K$ is split.
Further, for local fields at least one of $K^\times/(K^\times)^nN_{E/K}(E^\times)$ or $H^2(\GalKbK, T_{E,\sigma}) $ is trivial.
\end{remark}

Finally, we must consider the group $H^1(\GalKbK,G)$. As $G$ is simply connect, this is only non-trivial at the real places of $K$.
In terms of real places, there are two cases to consider.
If $E=\C$, then $G$ is the automorphism group of a Hermitian space and $H^1(\GalKbK,G)$ classifies the Hermitian forms of the same discriminant as that defining $G$, as above, over $\RR$ this means it classifies the signature.
When $E=\RR$ then $G$ is the automorphism group of a free module over a division algebra, there are no forms of these hence $H^1(\GalKbK,G)$ is trivial.

TODO --- Find reference for this ----

@@

One may check that $G$ determines $E$, $A$, $\delta$ and the signatures uniquely when $n>1$. Indeed, we may check this locally.
When $G$ splits $E=K\times K$, when $E$ is quasi-split then $E$ is the unique quadratic extension over which $G$ splits.
For places of $K$ which split in $E$, $G$ is precisely the norm one elements of the division algebra $A$, this uniquely determines $A$.
For places which do not splite, $G$ is the special unitary group of a Hermitian form of discriminant $\delta$.
The Hermitian form preserved by a special unitary group is unique up to scaling, $\delta$ respectively the signatures are the invariants of such forms.

\begin{remark}
In the case of $n=1$, we may perform the same constructions, however the group does not uniquely determine invariants $E$, $A$, $\delta$, and signatures (if we fix $E$, the others are uniquely determined as above).
Indeed, there is no outer automorphism group, so the choice of $E$ is no longer an invariant.
Exceptional isomorphisms of groups tell us that for $n=1$ the group $G$ is the spin group of a rank $3$ quadratic space.
In odd rank the discriminant does not impact the isomorphism type of the orthogonal group.
The discriminant of the quadratic space does still play a role in defining the Clifford algebra, in the same way it plays a role in defining a unitary group. The Witt invariant of the quadratic space is still an invariant and determines the algebra $A$.
